Fast equivalent layer technique for gravity data processing with Block-Toeplitz Toeplitz-Block matrix systems

### 2019-07-16 manuscript/toeplitz_manuscript.tex

\documentclass[paper]{geophysics}
%\documentclass[paper,revised]{geophysics}
\usepackage{listings}
% An example of defining macros
\newcommand{\rs}[1]{\mathstrut\mbox{\scriptsize\rm #1}}
\newcommand{\rr}[1]{\mbox{\rm #1}}

@@

\begin{abstract}
We present a new approach of the fast equivalent layer technique for gravity data processing, modifying \cite{siqueira2017fast}'s work, with the potential to use very large datasets at low computational cost. Taking advantage of the properties related to the symmetric Block-Toeplitz Toeplitz-block (BTTB) and Block-Circulant Circulant-Block (BCCB) matrices, that raises when regular grids of observation points and equivalent sources are used with the equivalent layer, we developed an algorithm which greatly reduces the number of flops and memory RAM necessary to complete the process of parameter estimative using this technique. The algorithm is based on the struture of symmetric BTTB matrices, where all its elements are comprised by the first row and can be embedded into a symmetric BCCB matrix, that also only needs its first row to be completed reconstructed. From the first column, the eigenvalues of BCCB matrices can be calculated using the Fast Fourier Transform, which can be used to readily compute matrix-vector products. Using examples, we demonstrate that even small and medium sized grids benefits from this approach and larger the dataset, faster and more efficient this method becomes compared to the usual. Synthetic tests using the equivalent layer with the method presented in this work evaluate this approach, and demonstrate satisfatory results for different gravity processing, as upward and downward continuation. Tests with real data from Caraj\'as, Brazil, shows its applicabillity and potential for field processing.
\end{abstract}

@@

\begin{equation}
\delta g(x, y, z) = c_{g} \, G \, \int\int\limits_{v}\int \rho(x^{\prime}, y^{\prime}, z^{\prime}) \frac{(z^{\prime} - z) \quad dx^{\prime}dy^{\prime}dz^{\prime}}{[(x - x^{\prime})^{2} + (y - y^{\prime})^{2} + (z - z^{\prime})^{2}]^{\frac{3}{2}}}\: ,
\label{eq:gravity-disturbance}
\end{equation}
where $c_{g} = 10^5$ is a constant transforming from $m/s^2$ to $mGal$, $G$ is the Newton's gravitational constant $(m^3/kg \, s^2)$ and $\rho(x^{\prime}, y^{\prime}, z^{\prime})$ is the density at the point $(x^{\prime}, y^{\prime}, z^{\prime})$ inside the volume $V$ of the source. This integral is defined in a Cartesian coordinate system with axis $x$ pointing to north, $y$ pointing to east and $z$ pointing downward.

@@

\begin{equation}
\delta g(x_i, y_i, z_i) = \frac{1}{2\pi}  \int\limits_{-\infty}^{+\infty} \int\limits_{-\infty}^{+\infty} \,
\frac{\delta g(x, y, z_0) (z_{0} - z_{i})}{\left[(x_{i} - x)^{2} +
	(y_{i} - y)^{2} +
	(z_{i} - z_{0})^{2} \right]^{\frac{3}{2}}} \, 
dx \, dy,
\label{eq:upward-continuation-integral}
\end{equation}
where, $z_0 > z_i$.
Equation \ref{eq:upward-continuation-integral} shows that the gravity disturbance $\delta g(x_{i}, y_{i}, z_{i})$ is a convolution between $\delta g(x, y, z_{0})$ and another harmonic function on the horizontal plane $z=z_0$.
Multiplying and dividing equation \ref{eq:upward-continuation-integral} by $G$ and discretizing the integral, we obtain:

\begin{equation}
\delta g(x_{i}, y_{i}, z_{i}) = \sum_{j=1}^{N} p_j a_{ij} \; ,
\label{eq:integral-sum}
\end{equation}
where $a_{ij}$ is given by:

\begin{equation}
a_{ij}= c_{g} \, G \, \frac{(z_{j} - z_{i})}{\left[(x_{i} - x_{j})^{2} +	(y_{i} - y_{j})^{2} +	(z_{i} - z_{j})^{2} \right]^{\frac{3}{2}}}
\label{eq:aij}
\end{equation}

\noindent and $p_j$ is the coefficient representing the physical property of the $j$-th equivalent source. In this case, the harmonic function $a_{ij}$ represents the vertical component of the gravitational attraction exerted, at the point $(x_{i}, y_{i}, z_{i})$, by a point mass located at the point $(x_{j}, y_{j}, z_{j})$, with mass $p_{j}$.
Equation \ref{eq:integral-sum} can be expressed in matrix form as:

\begin{equation}
\mathbf{d}(\mathbf{p}) = \mathbf{A} \mathbf{p} \: ,
\label{eq:predicted-data-vector}
\end{equation}
where $\mathbf{d}(\mathbf{p})$ is an $N \times 1$ vector, whose i-th element is the predicted gravity disturbance $\delta g(x_{i}, y_{i}, z_{i})$, $\mathbf{p}$ is an $N \times 1$ parameter vector,  whose j-th element is the coefficient $p_{j}$ representing the physical property of the $j$-th equivalent source and $\mathbf{A}$ is  an $N \times N$ sensibility matrix, where each element $a_{ij}$ is given by equation 4 and defines the field produced by  the j-th equivalent source at the point $(x_{i}, y_{i}, z_{i})$.
The solution of the parameters \textbf{p} can be solved by a linear inversion minimizing the function:

\begin{equation}
\Psi(\mathbf{p}) = \theta_g(\mathbf{p}) + \mu \, \theta_m(\mathbf{p}) \: ,
\label{eq:goal-function}
\end{equation}
where $\theta_g(\mathbf{p})$ is the misfit function given by:

\begin{equation}
\theta_g(\mathbf{p}) = ||\mathbf{d}^{o}-\mathbf{d(p)}||_2^2 \: ,
\label{eq:goal-function_d}
\end{equation}
which is the euclidian norm of the residual between the observed data $\mathbf{d}^{o}$ and the predicted data $\mathbf{d(p)}$.

@@

\begin{equation}
\theta_m(\mathbf{p}) = ||\mathbf{p}||_2^2 \: ,
\label{eq:tikhonov-function}
\end{equation}
which is the euclidian norm of the parameters $\mathbf{p}$. The variable $\mu$ is a real positive number regularizing the parameter.

Derivating equation \ref{eq:goal-function} in relation to $\mathbf{p}$ and making equal to zero, it is possible to estimate the parameters:

@@

\begin{equation}
\mathbf{p}^0 = \tilde{\mathbf{A}}^{-1} \mathbf{d}^{o} \: ,
\label{eq:initial_m}
\end{equation}
where $\mathbf{d}^o$ is the gravity data and $\tilde{\mathbf{A}}^{-1} = \Delta s/(2 \pi \, G \, c_g)$ with $\Delta s$ being an element of area located at the vertical coordinate $z_i$ and centered at the horizontal coordinates $(x_i,y_i), \, i = 1,..., N$. At each $k$-th iteration a mass correction vector $\mathbf{\Delta m^k}$ for every source is calculated by minimizing the function:

\begin{equation}
\phi(\mathbf{\Delta p^k}) = ||\mathbf{d}^{o} - \mathbf{A}\hat{\mathbf{p}}^k - \tilde{\mathbf{A}}\mathbf{\Delta p^k}||_2^2
\label{eq:goal-function_fast}
\end{equation}

\noindent and the mass distribution of the equivalent sources is updated as:

\begin{equation}
\mathbf{\Delta p^{k+1}} = \mathbf{p^{k}} + \mathbf{\Delta p^{k}} \: .
\label{eq:update_m}
\end{equation}

\lstset{language=Python}
\lstset{frame=lines}
\lstset{caption={A \textit{Python} algorithm for \cite{siqueira2017fast}'s fast equivalent layer.}}
\lstset{label={lst:code_direct}}
\lstset{basicstyle=\footnotesize}
\begin{lstlisting}
diagonal_A = G*(10**5)*2*pi/(dx*dy)
for i in range (itmax):
	res = (data - A.dot(p))
	delta_p = res/diagonal_A
	p += delta_p
\end{lstlisting}

Each iteration of this algorithm, the matrix-vector product $\tensor{A} \mathbf{p}$ must be calculated to get a new residual $\mathbf{d^0} - \tensor{A} \mathbf{p}$. While it is true that for a small number of observation points this represents no computational effort, for very large data sets it is costful and can be overwhelming in terms of memory RAM to maintain such operation.

\subsection{Structure of sensibility matrix A}
If a gravity disturbance  data $\delta g(x_i, y_i, z_i)$ (equation \ref{eq:integral-sum}) is placed on a regular grid at the constant vertical coordinate $z_{i} = z_{1}$, $i = 1, \dots, N$, and there is an equivalent source located directly below each point of this grid at a constant depth $z_{0}$, the elements $a_{ij}$ (equation \ref{eq:aij}) can be rewritten as follows:

\begin{equation}
a_{ij} = c_{g} G 
\frac{\Delta z}{\left[(r_{ij})^{2} + (\Delta z)^{2}\right]^{\frac{3}{2}}} \: ,
\label{eq:a_ij_r_ij}
\end{equation}
where $\Delta z = z_{0} - z_{1}$ and $r_{ij} = \sqrt{(x_{i} - x_{j})^{2} + (y_{i} - y_{j})^{2}}$ represents the  relative horizontal distance between the $i$-th observation point and the $j$-th equivalent source. It is worth noting that: (i) $r_{ij} = r_{ji}$ for any pair $ij$ and (ii) $r_{ij}$ computed for different pairs $ij$ may assume the same value. As a consequence of these two properties, the matrix $\mathbf{A}$ (equation \ref{eq:predicted-data-vector}) has the structure of a symmetric Toeplitz by blocks, where each block is also a symmetric Toeplitz matrix (BTTB) \cite[]{chan2007introduction, golub2013}. 
For example, consider a regular grid of $N_{x} \times N_{y}$ points, where $N_{x} = 4$ and $N_{y} = 3$, comprising a total number of $N = N_{x} \times N_{y} = 12$ points. In this case, matrix $\mathbf{A}$ is $N_{x} \times N_{x}$ grid of matrices given by:

\begin{equation}
\tensor{A} =
\pmatrix{
	\tensor{A_{0}} & \tensor{A_{1}} & \tensor{A_{2}} & \tensor{A_{3}} \cr
	\tensor{A_{1}} & \tensor{A_{0}} & \tensor{A_{1}} & \tensor{A_{2}} \cr
	\tensor{A_{2}} & \tensor{A_{1}} & \tensor{A_{0}} & \tensor{A_{1}} \cr
	\tensor{A_{3}} & \tensor{A_{2}} & \tensor{A_{1}} & \tensor{A_{0}}
},
\label{eq:toeplitz_block_T}
\end{equation}
where each block is a $N_{y} \times N_{y}$ matrix given by:

\begin{equation}
\tensor{A}_{l} =
\pmatrix{
	a_{0}^{\ell} & a_{1}^{\ell} & a_{2}^{\ell} \cr
	a_{1}^{\ell} & a_{0}^{\ell} & a_{1}^{\ell} \cr
	a_{2}^{\ell} & a_{1}^{\ell} & a_{0}^{\ell}
},
\label{eq:toeplitz_element}
\end{equation}
where the elements $a^{l}_{k}$, $l = 0, \dots, N_{x} - 1$, $k = 0, \dots, N_{y} - 1$ are computed with equation \ref{eq:a_ij_r_ij}. In this case, the complete matrix $\mathbf{A}$ can be obtained by computing only its first row or column.

\subsection{BCCB matrix-vector product}
As previous discussed in the  Fast equivalent layer technique, the matrix-vector product accounts for most of the computational cost (Listing \ref{lst:code_direct}). When large data sets are used, this operation can take some time and even be prohibited by memory RAM shortage. In order to lessen this problem we transform the BTTB matrix $\mathbf{A}$ into a Block-Circulating Circulating-Block (BCCB) matrix $\mathbf{C}$ and use its eigenvalues to carry the product of $\mathbf{A}$ and an arbitrary vector $\mathbf{p}$. This strategy has been successfully applied by Zhang and Wong (2015) and Zhang et al. (2016) for optmizing the computational cost of 3D gravity inversion and downward continuation of potential field, respectively. Here, we use this strategy for improving the computational efficiency of the Fast equivalent layer method proposed by Siqueira et al. (2017).

Following the example of our BTTB matrix $\mathbf{A}$ (equation \ref{eq:toeplitz_block_T}), its transformation into a BCCB matrix $\mathbf{C}$ is given by:

\begin{equation}
\tensor{C} =
\pmatrix{
	\tensor{C_{0}} & \tensor{C_{1}} & \tensor{C_{2}} & \tensor{C_{3}} & \tensor{0} & \tensor{C_{3}} & \tensor{C_{2}} & \tensor{C_{1}} \cr
	\tensor{C_{1}} & \tensor{C_{0}} & \tensor{C_{1}} & \tensor{C_{2}} & \tensor{C_{3}} & \tensor{0} & \tensor{C_{3}} & \tensor{C_{2}} \cr
	\tensor{C_{2}} & \tensor{C_{1}} & \tensor{C_{0}} & \tensor{C_{1}} & \tensor{C_{2}} & \tensor{C_{3}} & \tensor{0} & \tensor{C_{3}} \cr
	\tensor{C_{3}} & \tensor{C_{2}} & \tensor{C_{1}} & \tensor{C_{0}} & \tensor{C_{1}} & \tensor{C_{2}} & \tensor{C_{3}} & \tensor{0} \cr
	\tensor{0} & \tensor{C_{3}} & \tensor{C_{2}} & \tensor{C_{1}} & \tensor{C_{0}} & \tensor{C_{1}} & \tensor{C_{2}} & \tensor{C_{3}} \cr
	\tensor{C_{3}} & \tensor{0} & \tensor{C_{3}} & \tensor{C_{2}} & \tensor{C_{1}} & \tensor{C_{0}} & \tensor{C_{1}} & \tensor{C_{2}} \cr
	\tensor{C_{2}} & \tensor{C_{3}} & \tensor{0} & \tensor{C_{3}} & \tensor{C_{2}} & \tensor{C_{1}} & \tensor{C_{0}} & \tensor{C_{1}} \cr
	\tensor{C_{1}} & \tensor{C_{2}} & \tensor{C_{3}} & \tensor{0} & \tensor{C_{3}} & \tensor{C_{2}} & \tensor{C_{1}} & \tensor{C_{0}}
},
\label{eq:circulant_elements}
\end{equation}
where each block $\mathbf{C}_{l}$ is:

\begin{equation}
\tensor{C}_{l} =
\pmatrix{
	\tensor{A}_{l} & \times \cr
	\times & \tensor{A}_{l}
} =
\pmatrix{
	a_{0}^{\ell} & a_{1}^{\ell} & a_{2}^{\ell} & 0^{\ell} & a_{2}^{\ell} & a_{1}^{\ell} \cr
	a_{1}^{\ell} & a_{0}^{\ell} & a_{1}^{\ell} & a_{2}^{\ell} & 0^{\ell} & a_{2}^{\ell} \cr
	a_{2}^{\ell} & a_{1}^{\ell} & a_{0}^{\ell} & a_{1}^{\ell} & a_{2}^{\ell} & 0^{\ell} \cr
	0^{\ell} & a_{2}^{\ell} & a_{1}^{\ell} & a_{0}^{\ell} & a_{1}^{\ell} & a_{2}^{\ell} \cr
	a_{2}^{\ell} & 0^{\ell} & a_{2}^{\ell} & a_{0}^{\ell} & a_{0}^{\ell} & a_{1}^{\ell} \cr
	a_{1}^{\ell} & a_{2}^{\ell} & 0^{\ell} & a_{2}^{\ell} & a_{1}^{\ell} & a_{0}^{\ell}
}.
\label{eq:toeplitz_circulant_element}
\end{equation}

Matrix $\mathbf{C}$ is a $4N_x N_y \times 4N_x N_y$ Block-Circulant formed by Circulant-Blocks matrix. This matrix is formed by a grid of $2N_x \times 2N_x$ blocks, where each block is a $2N_y \times 2N_y$ matrix.

The product $\mathbf{d} = \mathbf{A}\mathbf{p}$ (equation \ref{eq:predicted-data-vector}), then becomes:

\begin{equation}
\mathbf{C} \mathbf{v} = \mathbf{q} \: ,
\label{eq:BCCB_vector_product}
\end{equation}
where $\mathbf{v}$ and $\mathbf{q}$ are $4N_x N_y \times 1$ vectors given by:

\begin{equation}
\mathbf{v} =
\pmatrix{
	\mathbf{v}_{0} \cr
	\mathbf{v}_{1} \cr
	\vdots \cr
	\mathbf{v}_{N_x- 1} \cr
	\mathbf{0}_{(2 N_x N_y)}
}
\end{equation}

\noindent and

\begin{equation}
\mathbf{q} =
\pmatrix{
	\mathbf{q}_{0} \cr
	\mathbf{q}_{1} \cr
	\vdots \cr
	\mathbf{q}_{N_x - 1} \cr
	\mathbf{0}_{(2 N_x N_y)}
} \: ,
\end{equation}
where $\mathbf{0}_{(2 N_x N_y)}$ is a $2 N_x N_y \times 1$ vetor of zeros, and $\mathbf{v}_{\ell}$ and $\mathbf{q}_{\ell}$, $\ell = 0, \dots, N_x-1$ are:

\begin{equation}
\mathbf{v}_{\ell} =
\pmatrix{
	\mathbf{p}_{\ell} \cr
	\mathbf{0}_{(N_y)}
}
\end{equation}

\noindent and

\begin{equation}
\mathbf{q}_{\ell} =
\pmatrix{
	\mathbf{d}_{\ell} \cr
	\mathbf{0}_{(N_y)}
} \: ,
\end{equation}
where $\mathbf{0}_{(N_y)}$ is a $N_y \times 1$ vetor of zeros.

By using the Kronecker product properties, the auxiliary matrix-vetor product can be rewritten as follows:

>>>>

\documentclass[paper]{geophysics}
%\documentclass[paper,revised]{geophysics}

% An example of defining macros
\newcommand{\rs}[1]{\mathstrut\mbox{\scriptsize\rm #1}}
\newcommand{\rr}[1]{\mbox{\rm #1}}

@@

\begin{abstract}
We present a new approach to the fast equivalent layer technique for gravity data processing, modifying \cite{siqueira2017fast}'s work, with the potential of using very large datasets at low computational cost. Taking advantage of the properties related to the symmetric Block-Toeplitz Toeplitz-block (BTTB) and Block-Circulant Circulant-Block (BCCB) matrices, that raises when regular grids of observation points and equivalent sources are used with the equivalent layer, we developed a algorithm which greatly reduces the number of flops and memory RAM necessary to complete the process of estimative using this technique. The algorithm is based on the struture of symmetric BTTB matrices, that all its elements are comprised by the first row and can be embedded into a symmetric BCCB matrix that also only needs its first row to be completed reconstructed. From this row, the eigenvalues of BCCB matrices are calculated using the Fast Fourier Transform, which can be used to readily compute matrix-vector products for this type of matrix. Using examples, we demonstrate that even small and medium sized grids benefits from this approach and larger the dataset, faster and more efficient this method becomes compared to the usual approach. Synthetic tests using the equivalent layer with the method presented in this work evaluate this approach, and demonstrate satisfatory results for different gravity processing, as upward and downward continuation. Tests with real data from Caraj\'as, Brazil, shows its applicabillity and potential for field processing.
\end{abstract}

@@

\begin{equation}
\delta g(x, y, z) = c_{g} \, G \, \int\int\limits_{v}\int \rho(x^{\prime}, y^{\prime}, z^{\prime}) \frac{(z^{\prime} - z) \quad dx^{\prime}dy^{\prime}dz^{\prime}}{[(x - x^{\prime})^{2} + (y - y^{\prime})^{2} + (z - z^{\prime})^{2}]^{\frac{3}{2}}}\: ,
\label{eq:gravity-disturbance}
\end{equation}

where $c_{g} = 10^5$ is a constant transforming from $m/s^2$ to $mGal$, $G$ is the Newton's gravitational constant $(m^3/kg \, s^2)$ and $\rho(x^{\prime}, y^{\prime}, z^{\prime})$ is the density at the point $(x^{\prime}, y^{\prime}, z^{\prime})$ inside the volume $V$ of the source. This integral is defined in a Cartesian coordinate system with axis $x$ pointing to north, $y$ pointing to east and $z$ pointing downward.

@@

\begin{equation}
\delta g(x_i, y_i, z_i) = \frac{1}{2\pi}  \int\limits_{-\infty}^{+\infty} \int\limits_{-\infty}^{+\infty} \,
\frac{\delta g(x, y, z_0) (z_{0} - z_{i})}{\left[(x_{i} - x)^{2} +
	(y_{i} - y)^{2} +
	(z_{i} - z_{0})^{2} \right]^{\frac{3}{2}}} \, 
dx \, dy,
\label{eq:upward-continuation-integral}
\end{equation}

where, $z_0 > z_i$.
Equation \ref{eq:upward-continuation-integral} shows that the gravity disturbance $\delta g(x_{i}, y_{i}, z_{i})$ is a convolution between $\delta g(x, y, z_{0})$ and another harmonic function on the horizontal plane $z=z_0$.
Multiplying and dividing equation \ref{eq:upward-continuation-integral} by $G$ and discretizing the integral, we obtain:

\begin{equation}
\delta g(x_{i}, y_{i}, z_{i}) = \sum_{j=1}^{M} p_j a_{ij}
\label{eq:integral-sum}
\end{equation}

where $a_{ij}$ is given by:

\begin{equation}
a_{ij}= c_{g} \, G \, \frac{(z_{j} - z_{i})}{\left[(x_{i} - x_{j})^{2} +	(y_{i} - y_{j})^{2} +	(z_{i} - z_{j})^{2} \right]^{\frac{3}{2}}},
\label{eq:aij}
\end{equation}

and $p_j$ is the coefficient representing the physical property of the $j$-th equivalent source. In this case, the harmonic function $a_{ij}$ represents the vertical component of the gravitational attraction exerted, at the point $(x_{i}, y_{i}, z_{i})$, by a point mass located at the point $(x_{j}, y_{j}, z_{j})$, with mass $p_{j}$.
Equation \ref{eq:integral-sum} can be expressed in matrix form as:

\begin{equation}
\mathbf{d}(\mathbf{p}) = \mathbf{A} \mathbf{p} \: ,
\label{eq:predicted-data-vector}
\end{equation}

where $\mathbf{d}(\mathbf{p})$ is an $N \times 1$ vector, whose i-th element is the predicted gravity disturbance $\delta g(x_{i}, y_{i}, z_{i})$, $\mathbf{p}$ is an $M \times 1$ parameter vector,  whose j-th element is the coefficient $p_{j}$ representing the physical property of the $j$-th equivalent source and $\mathbf{A}$ is  an $N \times M$ sensibility matrix, where each element $a_{ij}$ is given by equation 4 and defines the field produced by  the j-th equivalent source at the point $(x_{i}, y_{i}, z_{i})$.
The solution of the parameters \textbf{p} can be solved by a linear inversion minimizing the function:

\begin{equation}
\Psi(\mathbf{p}) = \theta_g(\mathbf{p}) + \mu \, \theta_m(\mathbf{p}) \: ,
\label{eq:goal-function}
\end{equation}

where $\theta_g(\mathbf{p})$ is the misfit function given by:

\begin{equation}
\theta_g(\mathbf{p}) = ||\mathbf{d^0}-\mathbf{d(p)}||_2^2 \: ,
\label{eq:goal-function_d}
\end{equation}

which is the euclidian norm of the residual between the observed data $\mathbf{d^0}$ and the predicted data $\mathbf{d(p)}$.

@@

\begin{equation}
\theta_m(\mathbf{p}) = ||\mathbf{p}||_2^2 \: ,
\label{eq:tikhonov-function}
\end{equation}

which is the euclidian norm of the parameters $\mathbf{p}$. The variable $\mu$ is a real positive number regularizing the parameter.

Deriving equation \ref{eq:goal-function} in relation to $\mathbf{p}$ and making equal to zero, it is possible to estimate the parameters:

@@

\begin{equation}
\mathbf{m}^0 = \tilde{\mathbf{A}}^{-1} \mathbf{d} \: ,
\label{eq:initial_m}
\end{equation}

where $\mathbf{d}^o$ is the gravity data and $\tilde{\mathbf{A}}^{-1} = \Delta s/(2 \pi \gamma)$ with $\Delta s$ being an element of area located at the vertical coordinate $z_i$ and centered at the horizontal coordinates $(x_i,y_i), \, i = 1,..., N$. At each $k$-th iteration a mass correction vector $\mathbf{\Delta m^k}$ for every source is calculated by minimizing the function:

\begin{equation}
\phi(\mathbf{\Delta m^k}) = ||\mathbf{d} - \mathbf{A}\hat{\mathbf{m}}^k - \tilde{\mathbf{A}}\mathbf{\Delta m^k}||_2^2 \: ,
\label{eq:goal-function_fast}
\end{equation}

and the mass distribution of the equivalent sources is updated as:

\begin{equation}
\mathbf{\Delta m^{k+1}} = \mathbf{m^{k}} + \mathbf{\Delta m^{k}} \: ,
\label{eq:update_m}
\end{equation}

Each iteration a matrix-vector product must be calculated to get a new residual. While it is true that for a small number of observation points this represents no computational effort, for very large data sets it is costful and can be overwhelming in terms of memory RAM to maintain such operation.

\subsection{Toeplitz matrix structure}
A Toeplitz matrix is one with constant diagonal, i.e., $\tensor{T} \in \tensor{R}^{N \times N} $ is Toeplitz if there exist scalars $r_{-N+1},\dots,r_{N-1}$ such that $r_{j-i}$ for all $i$ and $j$ \cite[]{golub2013,press2007numerical}.
Only the first row and the first column of  $\tensor{T}$ needs to be known in order to construct the whole matrix.

\begin{equation}
\tensor{T} =
\pmatrix{
r_{0} & r_{1} & \cdots & \cdots & r_{N-1} \cr
r_{-1} & r_{0} & r_{1} & \cdots & \vdots \cr
\vdots & r_{-1} & r_{0} & \ddots & \vdots \cr
\vdots & \cdots & \ddots & \ddots & r_{1} \cr
r_{1-N} & \cdots & \cdots &  r_{-1} & r_{0}
}.
\label{eq:toeplitz_T}
\end{equation}

\subsection{Structure of sensibility matrix A}
If a gravity disturbance  data $\delta g(x_i, y_i, z_i)$ (equation \ref{eq:integral-sum}) is placed on a regular grid, the matrix $\mathbf{A}$ (equation \ref{eq:predicted-data-vector}) has the structure of a symmetric Toeplitz by blocks, where each block is also a symmetric Toeplitz matrix (BTTB) \cite[]{chan2007introduction}. 
For example, consider a regular grid of $N_{x} \times N_{y}$ points, where $N_{x} = 4$ and $N_{y} = 3$, comprising a total number of $N = N_{x} \, N_{y} = 12$ points.  Consider also that there is an equivalent source located directly below each point of this grid. In this case, matrix $\mathbf{A}$ is $N \times N$ formed by a $N_{x} \times N_{x}$ grid of symmetric Toeplitz matrices, and each block is also a symmetric Toeplitz matrix with $N_{y} \times N_{y}$ elements:

\begin{equation}
\tensor{A} =
\pmatrix{
	\tensor{A_{0}} & \tensor{A_{1}} & \tensor{A_{2}} & \tensor{A_{3}} \cr
	\tensor{A_{1}} & \tensor{A_{0}} & \tensor{A_{1}} & \tensor{A_{2}} \cr
	\tensor{A_{2}} & \tensor{A_{1}} & \tensor{A_{0}} & \tensor{A_{1}} \cr
	\tensor{A_{3}} & \tensor{A_{2}} & \tensor{A_{1}} & \tensor{A_{0}}
},
\label{eq:toeplitz_block_T}
\end{equation}

the whole matrix is formed by $4 \times 4$ blocks, where each block is a $3 \times 3$ matrix with elements given by equation \ref{eq:aij}:

\begin{equation}
\tensor{A}_{l} =
\pmatrix{
	a_{0} & a_{1} & a_{2} \cr
	a_{1} & a_{0} & a_{1} \cr
	a_{2} & a_{1} & a_{0}
}.
\label{eq:toeplitz_element}
\end{equation}

Note that the case where matrix $\mathbf{A}$ is a symmetric BTTB with symmetric blocks, only its first row must be calculated.

\subsection{Circulant matrix structure}
A circulant matrix is a special case of Toeplitz matrices where each row is a cyclic shift of the row above it as $c_{-k} = c_{n-k}$ for $1 \le k \le n-1$:

\begin{equation}
\tensor{C} =
\pmatrix{
	c_{0} & c_{n-1} & \cdots & c_{2} & c_{1} \cr
	c_{1} & c_{0} & c_{n-1} & \cdots & c_{2} \cr
	\vdots & c_{1} & c_{0} & \ddots & \vdots \cr
	c_{n-2} & \cdots & \ddots & \ddots & c_{n-1} \cr
	c_{n-1}  & c_{n-2} & \cdots & c_{1} & c_{0}
}.
\label{eq:circulant_elements}
\end{equation}

Equally a symmetric Toeplitz, the first row of $\tensor{C}_n$ comprises all the different elements of the matrix.
An interesting recourse is that any $N \times N$ Toeplitz matrix can be embedded into a $2N \times 2N$ circulant matrix. For example, the Toeplitz matrix $\tensor{A}_{l}$ (equation \ref{eq:toeplitz_element}) is part of a circulant matrix:

\begin{equation}
\tensor{C}_{l} =
\pmatrix{
	\tensor{A}_{l} & \times \cr
	\times & \tensor{A}_{l}
} =
\pmatrix{
	a_{0} & a_{1} & a_{2} & 0 & a_{2} & a_{1} \cr
	a_{1} & a_{0} & a_{1} & a_{2} & 0 & a_{2} \cr
	a_{2} & a_{1} & a_{0} & a_{1} & a_{2} & 0 \cr
	0 & a_{2} & a_{1} & a_{0} & a_{1} & a_{2} \cr
	a_{2} & 0 & a_{2} & a_{0} & a_{0} & a_{1} \cr
	a_{1} & a_{2} & 0 & a_{2} & a_{1} & a_{0}
}.
\label{eq:toeplitz_circulant_element}
\end{equation}

This recourse can be extended to a BTTB matrix embedding it into a Block-Circulant Circulant-Block (BCCB), where each block is a circulant matrix as shown in equation \ref{eq:toeplitz_circulant_element} and the whole matrix is block circulant. Thus, a $N \times N$ BTTB matrix can be embedded into a $4N \times 4N$ BCCB matrix.

Of many interesting properties of circulant matrix, one in particular will be very useful: the eigenvalues of a circulant matrix comprise the discrete Fourier transform (DFT) of the first column of this matrix \cite[]{grenander1984szeg}. Thus, the eigenvalues of a $\tensor{C}$ circulant matrix can be calculated by the fast Fourier transform (FFT) as:

\begin{equation}
\Lambda_N = \sum_{j=0}^{N-1} c_j e^{\frac{2\pi \mathrm{i} jk}{N}}, \quad \mathrm{i} \equiv \sqrt{-1}
\label{eq:eigenvalues}
\end{equation}

where $j,k = 0,1,...,N-1$ are the entries of the matrix $\tensor{C}$.

\subsection{BCCB matrix-vector product}
As previous discussed in the  Fast equivalent layer technique, the matrix-vector product accounts for most of the computational cost. When large data sets are used, this operation can take some time and even be prohibited by memory RAM shortage . In order to lessen this problem we propose the use of the Circulant embedding of a Block-Toeplitz Toeplitz-Block into a Block-Circulating Circulating-Block matrix and computing by FFT the eigenvalues of this matrix to carry the multiplication with the parameter vector \cite[]{strang1986introduction, olkin1986linear}. Following the notation presented by \cite{chan2007introduction} the multiplication of a Toeplitz matrix embedded into a circulant matrix and an N-dimensional vector $\mathbf{p}$ is:

\begin{equation}
\pmatrix{
\tensor{T} & \times \cr
\times & \tensor{T}
}
\pmatrix{
\mathbf{p} \cr
\tensor{0}
} =
\pmatrix{
	\tensor{T} \mathbf{p} \cr
	\dagger 
},
\label{eq:toeplitz_circulant_product}
\end{equation}

where $\pmatrix{\tensor{T} & \times \cr	\times & \tensor{T}}$ is a $2N \times 2N$ matrix (equation \ref{eq:toeplitz_circulant_element}), $\tensor{0}$ is an $N$-dimensional vector of zeros and $\dagger$ will be drop out. Following the property presented in the previous section, one can take the FFT of the first column of this new circulant matrix and make the multiplication with the new zero-padded vector.

For a BTTB matrix-vector product, each Toeplitz block becomes a circulant block as in equation \ref{eq:toeplitz_circulant_product} and its correpondent vector is padded with zeros. The whole matrix then becomes a Block-Circulant, again padding the vector with zeros. As an example, the product between $N \times N$ matrix $\tensor{A}$ (equation \ref{eq:toeplitz_block_T}) and the $N$-dimensional vector $\mathbf{p}$ becomes:

\begin{equation}
\pmatrix{
	\tensor{C}_0 & \tensor{C}_1 & \tensor{C}_2 & \tensor{0}_b & \tensor{C}_2 & \tensor{C}_1 & \cr
	\tensor{C}_1 & \tensor{C}_0 & \tensor{C}_1 & \tensor{C}_2 & \tensor{0}_b & \tensor{C}_2 \cr
	\tensor{C}_2 & \tensor{C}_1 & \tensor{C}_0 & \tensor{C}_1 & \tensor{C}_2 & \tensor{0}_b \cr
	\tensor{0}_b & \tensor{C}_2 & \tensor{C}_1 & \tensor{C}_0 & \tensor{C}_1 & \tensor{C}_2 \cr
	\tensor{C}_2 & \tensor{0}_b & \tensor{C}_2 &\tensor{C}_1 & \tensor{C}_0 & \tensor{C}_1 \cr
	\tensor{C}_1 & \tensor{C}_2 & \tensor{0}_b & \tensor{C}_2 & \tensor{C}_1 & \tensor{C}_0
}
\pmatrix{
	\mathbf{q} \cr
	\tensor{0}
} =
\pmatrix{
	\tensor{C} \mathbf{q} \cr
	\dagger
},
\label{eq:BCCB_vector_product}
\end{equation}

where the matrix BCCB has $4N \times 4N$ dimension, each block $\tensor{C}_l$ is a circulant matrix defined in \ref{eq:toeplitz_circulant_element} and $\tensor{0}_b$ is a matrix of zeros with the same dimension of $\tensor{C}_l$. The new vector is $4N$, with $\mathbf{q}$ being a vector $2N$ padded with zeros in correct place, correspondent with the block sizes, and $\tensor{0}$ is a vector of zeros also $2N$. Despite this increase in dimension, only the first column of the matrix needs to be calculated. As before, the FFT will be applied to compute the eigenvalues of this new circulant matrix (the FFT of the first column of the matrix obtained in equation \ref{eq:BCCB_vector_product}) and also apply on the new parameter vector $\pmatrix{\mathbf{q} \cr \tensor{0}}$. To complete the process the inverse FFT must be applied:

\begin{equation}
\pmatrix{\tensor{C} \mathbf{q}\cr\dagger} = \mathcal{F}^{-1}[\Lambda_N * \mathcal{F}(\pmatrix{\mathbf{q} \cr \tensor{0}})],
\label{eq:fft_product}
\end{equation}

where $\Lambda_N$ is the eigenvalues of matrix BCCB defined in equation \ref{eq:eigenvalues}. By dropping out $\dagger$ we have the vector $\tensor{C}\mathbf{q}$ with $2N$ dimension. By once more dropping the correct leftovers correspondent of each block, the matrix-vector product $\tensor{A}\mathbf{p}$ is complete.

\subsection{Computation performance}
Equation \ref{eq:goal-function_fast} now can be calculated with the approach presented in the previous section. In a normal procedure of the fast equivalent layer, at each iteration a full matrix $\tensor{A}$ (equation \ref{eq:predicted-data-vector}) is multiplied by the estimated mass distribution parameter vector $\hat{\mathbf{m}}^k$. As pointed in \cite{siqueira2017fast} the number of flops (floating-point operations) necessary to estimate the $N$-dimensional parameter vector inside the iteration loop is:
